IPv6 over Low-Power Wireless Personal Area Networks\\ MSc. Thesis Proposal

### 2009-12-07 doc/msc-proposal/msc-proposal.tex
proposal: header compression

\newpage
\tableofcontents
%\listoffigures 
\newpage

%%
%	Introduction
%%
\chapter{Introduction}
* There are a lot of applications where we need many small devices + examples; why it's important;
* LoWPAN\\
* The purpose of the proposed work will be to investigate smth. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
* As preparation for this work, several related areas have been studied in more
depth and the collected background information; Structure of the proposal.\\

\chapter{IEEE 802.15.4}
The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks. The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes). Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.

@@

%%
%	Transmission of IPv6 Packets
%%
\chapter{IPv6 over IEEE 802.15.4}
A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising typically low cost and low power devices that conform to the IEEE 802.15.4-2003 standard. The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other. On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be specified that would provide end-to-end packet delivery including routing through intermediate hosts.

The Internet Protocol (IP) is the most widely deployed network layer protocol. Internet Protocol Version 4 (IPv4) is still the dominant protocol of the Internet, although the successor, Internet Protocol Version 6 (IPv6) \cite{rfc2460} is being deployed actively worldwide. The application of IP technologies to LoWPANs would provide a number of benefits. IP-based technologies are well-known and proven which allows to use the existing infrastructure. IP-based devices can be easily connected to other IP-based networks.

>>>>

\newpage
\tableofcontents
\listoffigures 
\newpage

%%
%	Introduction
%%
\chapter{Introduction}
* There are a lot of applications where we need many small devices + examples; why it's important;
* LoWPAN\\
* The purpose of the proposed work will be to investigate something. The proposal describes problems associated with enabling IP communication with devices in a LoWPAN.\\
* As preparation for this work, several related areas have been studied in more
depth and the collected background information; Structure of the proposal.\\

\chapter{IEEE 802.15.4}\label{ch:ieee802.15.4}
The IEEE 802.15.4 standard \cite{ieee802.15.4} specifies the physical layer and media access control for low-rate wireless personal area networks. The standard intends to offer the fundamental lower network layers of a type of wireless personal area network which focuses on low-cost, low-speed communication between tiny devices (motes). Such networks are typically limited to a personal operating space (POS) of up to 10 meters with a transfer rate of 250 kbit/s and involve little or no infrastructure.

@@

%%
%	Transmission of IPv6 Packets
%%
\chapter{6LoWPAN Networks}
A low-power wireless personal area network (LoWPAN) is a simple low throughput wireless network comprising low cost and low power devices that conform to the IEEE 802.15.4-2003 standard. The IEEE 802.15.4-2003 standard specifies node-to-node frame delivery between the devices which are within reachable distance from each other. On top of IEEE 802.15.4 a network layer (layer 3 of the seven-layer OSI model) has to be defined that would provide end-to-end packet delivery including routing through intermediate hosts.

%The Internet Protocol (IP) is the most widely deployed network layer protocol. Internet Protocol Version 4 (IPv4) is still the dominant protocol of the Internet, although the successor, Internet Protocol Version 6 (IPv6) \cite{rfc2460} is being deployed actively worldwide. The application of IP technologies to LoWPANs would provide a number of benefits. IP-based technologies are well-known and proven which allows to use the existing infrastructure. IP-based devices can be easily connected to other IP-based networks. Today there exist a lot of ready-to-use tools for diagnostics and management of IP networks.

6lowpan (IPv6 over Low power Wireless Personal Area Networks) is a working group within the IETF concerned with the specification of mechanisms to allow IPv6 packets to be sent to and received from over IEEE 802.15.4 based networks. The working group has already completed two documents: \cite{rfc4919} that documents and discusses the problem statement and \cite{rfc4944} that defines the format for the adaptation between IPv6 and 802.15.4. The latter document describes the frame format for transmission of IPv6 packets and the method of forming IPv6 link-local addresses and statelessly autoconfigured addresses on IEEE 802.15.4 networks.

As described in Chapter \ref{ch:ieee802.15.4}, IEEE 802.15.4 defines four types of frames: beacon frames, MAC command frames, acknowledgement frames, and data frames.  IPv6 packets are carried on data frames.  Data frames may optionally request acknowledgments to aid link-layer recovery.

\section{Addressing}
IEEE 802.15.4 defines addresses of two types: IEEE 64-bit extended addresses or 16-bit short addresses unique within the PAN. Both types are supported by 6LoWPAN.

6LoWPAN supports stateless address autoconfiguration, which allows to obtain the Interface Identifier \cite{rfc4291} for an IEEE 802.15.4 interface  based on the EUI-64 identifier \cite{eui64} \cite{rfc2464} assigned to the IEEE802.15.4 device. Even though all 802.15.4 devices have an EUI-64 address, 16-bit short addresses also can be used for address autoconfiguration.  In this case, a pseudo 48-bit address is formed by concatenating 16 zero bits to the 16-bit PAN ID, the resulting 32 bits are concatenated with the 16-bit short address and the interface identifier is formed from this 48-bit address:
\begin{center}\texttt{16\_bit\_PAN\_ID:16\_zero\_bits:16\_bit\_short\_address}\end{center}

The IPv6 link-local address for an IEEE 802.15.4 interface is formed by appending the Interface Identifier to the prefix \texttt{FE80::/64}.

Packets with a multicast IPv6 destination address are sent to the 16-bit 802.15.4 address obtained by concatenating the 3-bit multicast prefix 101, the last 5 bits in the 15-th octet and the whole 16-th octet of the multicast IPv6 address.

\section{Adaptation Layer}\label{sec:adapt.layer}
The IPv6 minimum Maximum Transmission Unit (MTU) size is defined as 1280 octets. As described in Chapter \ref{ch:ieee802.15.4}, the maximum IEEE 802.15.4 frame size is limited to 127 octets.  Taking into account the maximum frame overhead of 25 octets, 102 octets are left at the media access control layer. Link-layer security imposes further overhead, which in the maximum case (21 octets for AES-CCM-128) leaves only 81 octets available, which is far below the minimum IPv6 packet size. Therefor, a fragmentation and reassembly adaptation layer is provided at the layer below IP.

Even though 802.15.4 networks are expected to commonly use mesh routing, the IEEE 802.15.4-2003 \cite{ieee802.15.4} specification does not define such capability. Mesh routing allows two devices  without direct reachability to communicate with each other. In such mesh scenarios, intermediate devices are used as forwarders towards the final destination. In order to support mesh routing, in addition to the hop-by-hop source and destination link-layer addresses the originator and final destination addresses need to be known. For this purpose a Mesh Addressing header is introduced.

IPv6 datagrams transported over IEEE 802.15.4 are prefixed by an encapsulation header stack. Each header in the header stack contains a header type followed by zero or more header fields.  There are four encapsulation headers defined: Mesh Addressing Header, Broadcast Header, Fragmentation Header and Dispatch Header (described in Sections \ref{subsec:mesh.header}, \ref{subsec:frag.header},\ref{subsec:broad.header} and \ref{subsec:dispatch.header} respectively). These headers must appear in the specified order. First three of them are optional, whereas the last one is required.

\subsection{Dispatch Header}\label{subsec:dispatch.header}
The dispatch type is defined by a zero bit as the first bit followed by a one bit. The dispatch header is shown in Figure \ref{fig:dispatch.header}.

\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
                     1                   2                   3
   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
  |0 1| Dispatch  |  type-specific header
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{Dispatch Type and Header}\label{fig:dispatch.header}
\end{figure}
The Dispatch value identifies the type of payload immediately following the Dispatch Header. This can  be, for example, a compressed or uncompressed IPv6 header. More details can be found in \cite{rfc4944}.

\subsection{Mesh addressing Header}\label{subsec:mesh.header}

The mesh header type is defined by a one bit and zero bit as the first two bits. The mesh header is shown in Figure \ref{fig:mesh.header}:
\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
                       1                   2                   3
   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
  |1 0|V|F|HopsLft| originator address, final address
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{Mesh Addressing Type and Header}\label{fig:mesh.header}
\end{figure}

The \texttt{V} and \texttt{F} bits indicate whether an IEEE extended 64-bit address (EUI-64) or a short 16-bit addresses are used for the originator and the final destination addresses respectively. The \texttt{HopsLft} field is used as a hop limit and decremented by every forwarding node. The left part of the mesh header specifies the originator and final destination link-layer addresses.

\subsection{Fragmentation Header}\label{subsec:frag.header}
As previously discussed, an entire IPv6 datagram may not fit within a single 802.15.4 frame. In such case it is broken into link fragments. The first link fragment contains the first fragment header shown in Figure \ref{fig:first.fragment}.
\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
                      1                   2                   3
  0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
  |1 1 0 0 0|    datagram_size    |         datagram_tag          |
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{First Fragment}\label{fig:first.fragment}
\end{figure}

The second and subsequent link fragments contains a fragmentation header that conforms to the format shown in Figure \ref{fig:sub.fragment}.

\begin{figure}[htp]
\centering
\begin{mylisting}
\begin{verbatim}
                     1                   2                   3
   1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5 6 7 8 9 0 1
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
  |1 1 1 0 0|    datagram_size    |         datagram_tag          |
  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
  |datagram_offset|
  +-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{Subsequent Fragments}\label{fig:sub.fragment}
\end{figure}

The first and subsequent fragment header types are specified by the binary sequences \texttt{11000} and \texttt{11100} respectively.  The  \texttt{datagram\_size} 11-bit field encodes the size of the entire IP packet before link-layer fragmentation.  The \texttt{datagram\_tag} field is the same for all link fragments of a datagram and incremented by the sender for successive fragmented datagrams. The \texttt{datagram\_offset} field is present only in the second and subsequent link fragments and specifies the offset of the fragment from the beginning of the payload datagram.

\subsection{Broadcast Header}\label{subsec:broad.header}
Broadcast is an additional mesh routing functionality. For that a broadcast header is defined the format of which is shown in Figure \ref{fig:broad.header}.
\begin{figure}[htp]
\begin{mylisting}
\begin{verbatim}
                                       1
                   0 1 2 3 4 5 6 7 8 9 0 1 2 3 4 5
                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
                  |0|1|0|1|0|0|0|0|Sequence Number|
                  +-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+-+
\end{verbatim}
\end{mylisting}
\caption{Broadcast Header
}\label{fig:broad.header}
\end{figure}

The broadcast header consists of a dispatch type \texttt{01010000} followed by a sequence number field, which is used to detect duplicate packets.
This field is incremented by the originator whenever it sends a new mesh broadcast or multicast packet.

\section{Header Compression}
As described in Section \ref{sec:adapt.layer}, 81 octets are left in a IEEE 802.15.4 frame for IPv6. The IPv6 header is 40 octets long which leaves only 41 octets for upper-layer protocols, like UDP.  The latter uses 8 octets in the header, which leaves 33 octets for application data. 
The fragmentation and reassembly layer, described in Section \ref{sec:adapt.layer}, will also use at least one additional octet, what makes header compression almost being unavoidable.

RFC 4944 \cite{rfc4944} defines stateless header compression mechanisms for IPv6 datagrams (LOWPAN\_HC1 and LOWPAN\_HC2) to reduce the relatively large IPv6 and UDP headers. However, LOWPAN\_HC1 and LOWPAN\_HC2 are insufficient for most practical uses of 6LoWPAN networks, since they are effective only for link-local unicast communication, where IPv6 addresses derived directly from IEEE 802.15.4 addresses. When communicating with devices external to the LoWPAN or in a route-over configuration where IP forwarding occurs within the LoWPAN, LOWPAN\_HC1 carries both IPv6 source and destination addresses in-line. Internet Draft document \cite{draft-hc-06} defines an encoding format, LOWPAN\_IPHC, for effective compression of unique local, global, and multicast IPv6 addresses based on shared contexts. In addition, it also defines an encoding format, LOWPAN\_NHC, for arbitrary next header compression.

\subsection{HC1 Header Encoding}
HC1 compression format does not use any shared context and compresses headers relying on information pertaining to the entire link.

\subsection{HC2 Header Encoding}

\subsection{IPHC Header Encoding}

\subsection{NHC Header Encoding}

\chapter{6LoWPAN Neighbor Discovery}
